---
title: "Intro to `ggplot`"
format: 
  revealjs:
    code-fold: false
filters:
  - webr
engine: knitr

---

# Preliminaries

## Announcements

- Exercise 02 due today
- No class Monday, September 21

## Today's topics

- Intro to ggplot
- As time permits
  - [Exercise 02](../exercises/ex-02-data-viz-outside-psych.qmd)
  - [Exercise 03](../exercises/ex-03-making-data.qmd)

# Intro to ggplot

## What is it?

- R package that implements Cleveland's "grammar of graphics"
  - Written by Hadley Wickham, CTO of Posit.co
- Map data to figure components
  - Position
  - Color, shape, size
- Add layers to figures

## Why 'code' your figures?

- Easier to reproduce
- Easier to edit
- Try out different styles, pick the best
- Be kind to your future (forgetful) self
- Makes programming more fun

## Installing ggplot

```r
install.packages("ggplot2")
```

## Let's make some data

```{r}
days <- factor(
  c("sun", "mon", "tue", "wed", "thu", "fri", "sat"),
  levels = c("sun", "mon", "tue", "wed", "thu", "fri", "sat"),
  ordered = TRUE
)
meals <- factor(
  c("breakfast", "lunch", "dinner", "snack", "dessert"),
  levels = c("breakfast", "lunch", "snack", "dinner", "dessert"),
  ordered = TRUE
)
meal_days <- rep(days, each = length(meals))
weekly_meals <- rep(meals, 7)

my_meals <- data.frame(day_of_wk = meal_days, 
                       meal = weekly_meals)
```

## Let's make some data

```{r}
ate_skipped <- sample(c("ate", "skipped"), size = 35, 
                 replace = TRUE, prob = c(.6, .4))
ate <- ate_skipped == "ate"
ate
```

## Let's make some data

```{r}
ratings <- sample(c(1, 2, 3, 4, 5), size = 35,
                 replace = TRUE)

ratings[!ate] <- NA
```

## Let's make some data

```{r}
my_meals$ate_skipped <- ate_skipped
my_meals$rating <- ratings
```

## How'd we do?

```{r}
my_meals
```

## Your turn...

::: {.question}
## How would you visualize this sort of data?
:::

::: {.fragment}
::: {.question}
## What questions would you want your visualization to answer?
:::
:::

## Meal ratings

```{r}
# Do this once each time you start work
library(ggplot2)
```

## Figure 0

```{r}
#| echo: true
#| output-location: slide
library(ggplot2)
ggplot(airquality, aes(Temp, Ozone)) + 
  geom_point() + 
  geom_smooth(method = "loess")
```

## Figure 1.0

```{r}
#| label: fig-barplot-ate_skipped
#| fig-cap: "A barplot of meals eaten and skipped."
#| echo: true
#| output-location: slide
#| code-line-numbers: "1|2|3|4"
my_meals |> 
  ggplot() + 
  aes(x = ate_skipped) + 
  geom_bar() 
```

## Figure 1.1

- Add a `fill` aesthetic to fill the bars with colors 

```{r}
#| label: fig-barplot-ate_skipped-fill
#| fig-cap: "A barplot of meals eaten and skipped."
#| echo: true
#| output-location: slide
#| code-line-numbers: "|3"
my_meals |>
  ggplot() +
  aes(x = ate_skipped, fill = ate_skipped) +
  geom_bar()
```

::: {.fragment}
- If we fill from `ate_skipped`
  - ggplot will map a color to each distinct value of `ate_skipped`
  - and add a legend, so we know what the colors represent
  
:::

## Figure 2

```{r}
#| label: fig-barplot-ratings
#| fig-cap: "A barplot of meal ratings"
ggplot(data = my_meals, 
       aes(x = rating)) +
  geom_bar()
```

## Figure 2

```{r}
#| label: fig-2
#| fig-cap: "A plot of ratings for each meal using  `geom_point`"
my_meals |>
ggplot() +
  aes(x = meal, y = rating) +
  geom_point()
```

## Figure 3

```{r}
#| label: fig-3

my_meals |>
ggplot() +
  aes(x = day_of_wk, y = rating) +
  geom_point()
```

## Basic ggplot components

- A data frame
- Function call to `ggplot()`
- Add layers to plot with `+` operator
- Add aesthetics (mappings between data and graphic elements) using `aes()` with x (and often y) mappings
- A `geom_` layer

# Wrap-up

## Next time

- Figure components
- Work session: Exercises 02 and 03

# Resources

## About

This talk was produced using [Quarto](https://quarto.org), using the [RStudio](https://posit.co/products/open-source/rstudio/)
Integrated Development Environment (IDE), version 2026.9.0.174.

The source files are in R and Quarto, then rendered to HTML using the
[revealJS](https://revealjs.com) framework.
The HTML slides are hosted in a [GitHub repo](https://github.com/psu-psychology/psych-490-data-viz-2026-fall) and served by GitHub pages:
<https://psu-psychology.github.io/https://github.com/psu-psychology/psych-490-data-viz-2026-fall/>

## References