---
title: "*ggplot2* tutorial II"
format:
  html:
    code-fold: false
---

## About

This page provides some activities for beginning users of the *ggplot2* [@ggplot2] package. It is the second of two tutorials.

## Get ready

1. Create a new project.

2. Customize the document
  - Change the title
  - Add an author field
  - Add format text
  
The full header should look like this:

::: {#fig-yaml-header}

```yaml
---
title: "psych-490-ggplot-2026-09-25"
author: "Your name"
format:
  html:
    self_contained: true
---
```

Basic header information for your document for this tutorial.
:::

3. Add a setup section 

Below the setup header, we're going to add an R code chunk like this:

::: {#fig-library-ggplot2}

````{{r}}
library(ggplot2)
````

Bring ggplot into your local computational environment.
:::

## Good habits

One of the first things data analysts should do when we get our hands on a new dataset is answer some basic questions about it, specifically the following:

- How many cases (data points) are in the dataset?
- How many variables?
- Which variables are nominal, which are ordinal, and which are continuous?

R makes it easy to answeer these questions for us.
We use the `summary()` and `dim()` (dimension) commands.

Add a section to your Quarto document for summary data about the *diamonds* dataset.
Then, add the following code chunk.

::: {#fig-summary-diamonds}
````{{r}}
summary(diamonds)

dim(diamonds)
````

R code to provide summary information about the *diamonds* dataset.
:::

Save your Quarto document and render it.

::: {#fig-summary-diamonds-output}

```{r}
#| echo: false
library(ggplot2)
summary(diamonds)

dim(diamonds)
```

Output from @fig-summary-diamonds
:::

Notice that the `summary()` function does what it says; it provides a statistical summary of the variables in the dataset.
For continuous variables, we get the minimum and maximum, the mean and median, and the 1st and 3rd quartile values. For discrete variables (`cut`, `color`, and `clarity`), we get informatoin about the number of cases at each level.

The `dim()` command tells us that there are 53,940 rows or individual cases and 10 columns or variables.

How do we know whether the discrete variables are ordered?

One way is to use the `unique()` command.
If the dataset has recorded that a variable is ordinal, this command will show us.

::: {#fig-diamonds-unique}

````{{r}}
with(diamonds, unique(cut))

with(diamonds, unique(color))

with(diamonds, unique(clarity))
````

Code to report the number of unique values in a variable and whether they have an intrinsic order.
:::

::: {#fig-diamonds-unique-output}

```{r}
#| echo: false
with(diamonds, unique(cut))

with(diamonds, unique(color))

with(diamonds, unique(clarity))
```

Output from @fig-diamonds-unique
:::

You might add this to your document so that you have a reasonably complete summary of the dataset before we visualize it.

::: {.callout-warning}
## Consistency alert

Gilmore usually advocates for a visualize first, analyze later approach to data exploration.
But here, we're generating statistical summaries before we've done any 'visualization.'
What gives?

Well, Gilmore should probably revise his advice to say *summarize* before you analyze.
:::

## Summarizing figures

A best practice is to generate summary figures for all the variables in your dataset.
What should the figures summarize, or said in the language of *ggplot*, what `geom_` should we  use? Histograms (`geom_histogram()`), boxplots (`geom_boxplot()`), and violin (`geom_violin()`) plots are great for continuous data, but what about nominal (ordered or unordered) data?

The following code chunk provides these for `carat` and `price`.

::: {#fig-summary-figures-carat-price}

````{{r}}
ggplot(data = diamonds) +
  aes(x = carat) +
  geom_histogram()
  
ggplot(data = diamonds) +
  aes(x = carat, y = "") +
  geom_violin() +
  geom_boxplot(alpha = .4) +
  ylab("")
````

Code to generate a histogram and combined boxplot/violin plot of the `carat` variable.
:::

Create a section of your Quarto document for these figures and make plots for the continuous variables.

## References
