---
title: "Anatomy of a Quarto document"
author: ["Rick Gilmore", "Santa Claus"]
date: today
format: 
  html:
    embed-resources: true
---

## About

This is a second-level header section.
The `About` label is preceded by two hashmarks (pound signs `##`).

Gilmore recommends adding a section like this to *every* document.
It's a gift to your future (forgetful) self.

## Setup

Gilmore likes to include a setup section.
The section includes a code chunk where he likes to add code needed to get to work.
If he finds that he needs another package, he should install the package in the R console then add the `library()` command here.

```{r}
library(ggplot2)
```

## Load data

## Make figures

### Figure 1

Your goals for Figure 1 should go here.
The R code to generate Figure 1 should go in the code chunk below.

```{r}

```

::: {.callout-important}
## Use a sensible topic/header hierarchy

Notice that the header for this figure has three hashmarks `###`.
That means it is a level 3 header, below the level 2 header called `Making figures`.

This isn't the only way to organize information for this report, but it is an easy one to follow.
:::

### Figure 2

Code for Figure 2 goes in the chunk below.

### Figure 3

Code for Figure 3 goes in the chunk below.

## Discussion and Conclusion

## Resources

Put your resources here.