---
title: "Student-generated surveys"
format: html
---

## About

This page provides links to surveys that students generated for [Exercise 03](../exercises/ex-03-making-data.qmd).

## Setup

Bring some required packages into the local environment.

```{r}
library(readr)
library(dplyr)
```

## Import data

Dr. Gilmore created a Google sheet to keep track of the data.

::: {.callout-note}
## Manual vs. automatic downloading

Typically, Dr. Gilmore would download the data using code.
Here he manually downloaded the file and saved it as a CSV under `private/csv`.

This avoids issues related to Google authentication, but requires an additional manual step.

:::

We read the data file into a data frame.

```{r}
surveys_df <- readr::read_csv("../private/csv/psych-490.007-2026-fall-surveys - Sheet1.csv", show_col_types = FALSE)
```

## Summarize

```{r}
summary(surveys_df)
```

There were *n*=`{r} dim(surveys_df)[1]` submissions.

*N*=`{r} sum(is.na(surveys_df$edit_url))` submissions did *not* include an edit link. Check [below](#editor-links) to see if you provided that link.

*N*=`{r} sum(is.na(surveys_df$response_url) & surveys_df$canvas_submission == "yes")` students submitted write-ups to Canvas but did not send the edit link to the survey to Dr. Gilmore.

## Links tables

### Responder links

Please respond to each survey.
The data you provide will help with [Exercise 05](../exercises/ex-05-tidy-data.qmd).

```{r}
surveys_df |>
  filter(!is.na(response_url) & canvas_submission == "yes") |>
  select(response_url) |>
  mutate(url = paste0("<", response_url, ">")) |>
  select(url) |>
  kableExtra::kable()
```

### Editor links

Please confirm that you *cannot* open these survey links, with one exception--The survey that *you* created.
If your survey is not in this list, then open your survey and add Dr. Gilmore (rog1@psu.edu) and Kalvin Garrah (cfh5558@psu.edu) as collaborators on the survey as indicated in the [Exercise 03](../exercises/ex-03-making-data.qmd) instructions.
This will let us help you with [Exercise 05](../exercises/ex-05-tidy-data.qmd).

```{r}
surveys_df |>
  filter(!is.na(edit_url) & canvas_submission == "yes") |>
  select(edit_url) |>
  mutate(url = paste0("<", edit_url, ">")) |>
  select(url) |>
  kableExtra::kable()
```
